The evolution of gene expression in the term placenta of viviparous mammals

### 2015-12-15 manuscript/manuscript.Rnw
use xspace
work on discussion of SSLP1

\newcommand{\smonkey}[0]{\textit{A. fusciceps}}
\newcommand{\cow}[0]{\textit{B. taurus}}
\newcommand{\dog}[0]{\textit{C. familiaris}}
\newcommand{\armadillo}[0]{\textit{D. novemcinctus}}
\newcommand{\horse}[0]{\textit{E. galili}}
\newcommand{\elephant}[0]{\textit{L. africana}}
\newcommand{\human}[0]{\textit{H. sapiens}}
\newcommand{\opossum}[0]{\textit{M. domestica}}
\newcommand{\mouse}[0]{\textit{M. musculus}}
\newcommand{\galili}[0]{\textit{N. galili}}
\newcommand{\sheep}[0]{\textit{O. aries}}
\newcommand{\bonobo}[0]{\textit{P. paniscus}}
\newcommand{\carmeli}[0]{\textit{S. carmeli}}
\newcommand{\boar}[0]{\textit{S. scrofa}}

@@

\subsection{Dog Placenta Transcriptome}

@@

The top 50 non-housekeeping expressed genes as measured by \ac{FPKM}
in the term placenta of \textit{Canis familiaris} are shown in
\cref{tab:top50_dog_placenta} with a list of all genes in Supplemental
\cref{tab:all_canis_familiaris_placenta}. The gene with the highest
expression %
(\ac{FPKM}=%
\Sexpr{gene.expression("canis familiaris","ENSCAFG00000002115")}) %
is the hormone relaxin (\textit{RLN2}). Other highly expressed genes
among the top 50 (\cref{tab:top50_dog_placenta}) include genes
involved in the cytoskeleton (\textit{TMSB10}, \textit{TMSB4X}),
anticoagulation (\textit{TFPI2}, \textit{PLA2G7}), innate immunity
(\textit{LCN2}, \textit{GM2A}), lipid metabolism (\textit{APOE}),
inhibition of matrix metalloproteinases (\textit{TIMP2},
\textit{TIMP3}), and growth factors (\textit{PGF}, \textit{EGFR},
\textit{HTRA1}). The gene \textit{LGALS1}, which shows evidence of
immunosupression during pregnancy
\cite{Than.ea2008:Emergenceofhormonaland}, is highly expressed %
(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","LGALS1")}). %
In addition, the gene \textit{SSLP1}, a gene expressed in the
placentas of \textit{Bos tarus} (%
\Sexpr{gene.expression("bos tarus","SSLP1")}, %
\cref{tab:top50_cow_placenta}) and \textit{Ovis Aries}
(\Sexpr{gene.expression("ovis aries","SSLP1")},
\cref{tab:top50_sheep_placenta}) but is not present in primates
\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}, is also
expressed at high levels
%
(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","SSLP1")}). %
There is no placental expression of the estrogen synthase gene
(CYP19A1) in dogs, however expression of HSD3B1 is relatively high
(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","HSD3B1")}).

@@

tree.id <-
    genes.to.tree[combined.fpkm[gene_short_name=="SSLP1" & species=="homo sapiens",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{Expression of prolactin receptor}
\label{fig:prolactin_receptor_evolution_tree}
\end{figure}

>>>>

\newcommand{\smonkey}[0]{\textit{A. fusciceps}\xspace}
\newcommand{\cow}[0]{\textit{B. taurus}\xspace}
\newcommand{\dog}[0]{\textit{C. familiaris}\xspace}
\newcommand{\armadillo}[0]{\textit{D. novemcinctus}\xspace}
\newcommand{\horse}[0]{\textit{E. galili}\xspace}
\newcommand{\elephant}[0]{\textit{L. africana}\xspace}
\newcommand{\human}[0]{\textit{H. sapiens}\xspace}
\newcommand{\opossum}[0]{\textit{M. domestica}\xspace}
\newcommand{\mouse}[0]{\textit{M. musculus}\xspace}
\newcommand{\galili}[0]{\textit{N. galili}\xspace}
\newcommand{\sheep}[0]{\textit{O. aries}\xspace}
\newcommand{\bonobo}[0]{\textit{P. paniscus}\xspace}
\newcommand{\carmeli}[0]{\textit{S. carmeli}\xspace}
\newcommand{\boar}[0]{\textit{S. scrofa}\xspace}

@@

\subsection{\dog Placenta Transcriptome}

@@

The top 50 non-housekeeping expressed genes as measured by \ac{FPKM}
in the term placenta of \dog are shown in
\cref{tab:top50_dog_placenta} with a list of all genes in Supplemental
\cref{tab:all_canis_familiaris_placenta}. The gene with the highest
expression %
(\ac{FPKM}=%
\Sexpr{gene.expression("canis familiaris","ENSCAFG00000002115")}) %
is the hormone relaxin (\textit{RLN2}). Other highly expressed genes
among the top 50 (\cref{tab:top50_dog_placenta}) include genes
involved in the cytoskeleton (\textit{TMSB10}, \textit{TMSB4X}),
anticoagulation (\textit{TFPI2}, \textit{PLA2G7}), innate immunity
(\textit{LCN2}, \textit{GM2A}), lipid metabolism (\textit{APOE}),
inhibition of matrix metalloproteinases (\textit{TIMP2},
\textit{TIMP3}), and growth factors (\textit{PGF}, \textit{EGFR},
\textit{HTRA1}). The gene \textit{LGALS1}, which shows evidence of
immunosupression during pregnancy
\cite{Than.ea2008:Emergenceofhormonaland}, is highly expressed %
(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","LGALS1")}). %
In addition, an ortholog of SSLP1 (ENSCAFG00000010638) a gene
expressed in the placentas of \cow (%
\Sexpr{gene.expression("bos tarus","SSLP1")}, %
\cref{tab:top50_cow_placenta}) and \sheep
(\Sexpr{gene.expression("ovis aries","SSLP1")},
\cref{tab:top50_sheep_placenta}) but is not present in primates
\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}
(\cref{fig:sslp1_evolution_tree}), is also expressed at high levels
%
(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","ENSCAFG00000010638")}). %
There is no placental expression of the estrogen synthase gene
(\textit{CYP19A1}) in \dog (~\ref{fig:estrogen_evolution_tree}), 
however expression of \textit{HSD3B1} is relatively high
(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","HSD3B1")}).

@@

tree.id <-
    genes.to.tree[combined.fpkm[gene_id=="ENSCAFG00000010638" & species=="canis familiaris",gene_id],tree]
gene.tree.table <- trees[[tree.id]]$table
gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
@ 
\caption{Expression of SSLP1}
\label{fig:sslp1_evolution_tree}
\end{figure}
